---
title: "Baseball Contracts"
author: "Ben Scartz, Jack Arbuckle"
date: "2023-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE}
library(httr)
library(jsonlite)
library(rvest)
library(jsonlite)
library(lubridate)
library(tidyverse)
library(baseballr)
```

# Create function to scrape contract data from spotrac.com

```{r}

get_contracts <- function(position) {

url <- paste0('https://www.spotrac.com/mlb/contracts/sort-value/',position,'/all-time/limit-2000/')

contracts_page <- read_html(url)
 contracts_df <- contracts_page %>%
  html_table() %>%
  .[[1]]
 
 contracts_df$name <-  contracts_df$Player %>%
   str_remove(pattern = '\\w+') %>%
   str_remove(pattern = 'Jr.') %>%
   str_extract(pattern = '[^\\s]+ [^\\s]+( [^\\s]+)?')
 
 contracts_df$sign_year <- contracts_df$Player %>%
   str_extract(pattern = '\\d{4}')
 
 contracts_df$end_year <- contracts_df$Player %>%
   str_extract(pattern = '\\d{4}-\\d{4}') %>%
   str_extract(pattern = '\\d{4}$')
 
 contracts_df <- contracts_df %>%
   select(name, `Signed Age`, Yrs, Value, AAV, 
          `Signing Bonus`, sign_year, end_year) %>%
   mutate(`Signed Age` = as.numeric(`Signed Age`))
 
 contracts_df <- contracts_df %>%
   rename(sign_age = `Signed Age`,
          length = Yrs,
          total_value = Value,
          aav = AAV,
          sign_bonus = `Signing Bonus`)
 
 contracts_df$total_value <- as.integer(str_remove_all(string = contracts_df$total_value,
                                                       pattern = '\\$|,'))
 
 contracts_df$aav <- as.integer(str_remove_all(string = contracts_df$aav, 
                                                pattern = '\\$|,'))
 
 contracts_df$sign_bonus <- as.integer(str_remove_all(string = contracts_df$sign_bonus, 
                                                pattern = '\\$|,'))
 
 options(scipen = 999)
 contracts_df$sign_year <- as.integer(contracts_df$sign_year)
 contracts_df$end_year <- as.integer(contracts_df$end_year)
 contracts_df$year1 <- contracts_df$sign_year - 1
 contracts_df$year2 <- contracts_df$sign_year - 2
 contracts_df$year3 <- contracts_df$sign_year - 3
 
 contracts_df <- contracts_df %>%
   filter(aav > 720000 & sign_year >= 2019)
 
 return(contracts_df)
}

contracts_c <- get_contracts('catcher')
head(contracts_c)
```

# Calculate statistics using MLB Statcast data

Load Statcast data
```{r}
#load("G:/My Drive/Baseball/R Projects/Data/Statcast2022.RData")
#load("G:/My Drive/Baseball/R Projects/Data/Statcast2021.RData")
#load("G:/My Drive/Baseball/R Projects/Data/Statcast2020.RData")
#load("G:/My Drive/Baseball/R Projects/Data/Statcast2019.RData")
#load("G:/My Drive/Baseball/R Projects/Data/Statcast2018.RData")

```

```{r}
load("Statcast2022.RData")
load("Statcast2021.RData")
load("Statcast2020.RData")
load("Statcast2019.RData")
load("Statcast2018.RData")
```

Create function to calculate batter stats for a season 

```{r}
get_stats <- function(year) {
# From Statcast data, calculate metrics for qualified players
Statcast <- get(paste0('Statcast',year))
# Get unique values from the 'batter' column of Statcast
unique_playerID <- unique(Statcast$batter)

# Create empty data frame for Metrics
Metrics <- data.frame()

# Define label vectors and counts
hits_vector <- c('single', 'double', 'triple', 'home_run')
on_base_vector <- c(hits_vector, 'walk', 'hit_by_pitch','catcher_interf')
at_bats_vector <- c('field_out', 'strikeout', 'grounded_into_double_play', 'fielders_choice', 'fielders_choice_out', 'triple_play', 'strikeout_double_play', 'double_play', 'field_error', 'force_out', hits_vector)
plate_appearances_vector <- c(at_bats_vector, 'sac_fly', 'walk', 'hit_by_pitch', 'sac_bunt', 'sac_bunt_double_play', 'sac_fly_double_play', 'catcher_interf')
swings_vector <- c('hit_into_play', 'swinging_strike_blocked', 'swinging_strike', 'foul_tip', 'foul')

# Loop through each unique player ID
for (player_id in unique_playerID) {
  # Filter Statcast data for the current player ID and game_type == 'R'
  filtered_data <- filter(Statcast, batter == player_id & game_type == 'R')
  
  # Define count labels
  walk_count <- sum(filtered_data$events == 'walk')
  strikeout_count <- sum(filtered_data$events == 'strikeout')
  hit_by_pitch_count <- sum(filtered_data$events == 'hit_by_pitch')
  barrel_count <- sum(filtered_data$barrel, na.rm = TRUE)
  home_run_count <- sum(filtered_data$events == 'home_run')
  swing_count <- sum(filtered_data$description %in% swings_vector)
  
  # Calculate desired Metrics for the current player ID
  
  # Calculate plate appearances (PA)
  PA <- sum(filtered_data$events %in% plate_appearances_vector)
  
   # Calculate Batting Average
  batting_avg <- sum(filtered_data$events %in% hits_vector) / sum(filtered_data$events %in% at_bats_vector)
  
  #Calculate OBP
  on_base_percentage <- sum(filtered_data$events %in% on_base_vector) / sum(filtered_data$events %in% plate_appearances_vector)
  
  #Calculate Slug%
  slug_percentage <- (sum(filtered_data$events == 'single') + 
                        2 * sum(filtered_data$events == 'double') +
                        3 * sum(filtered_data$events == 'triple') +
                        4 * sum(filtered_data$events == 'home_run')) /
    sum(filtered_data$events %in% at_bats_vector)
  
  #Calculate OPS
  ops <- slug_percentage + on_base_percentage
  
  # Calculate HR
  hr <- home_run_count
  
  # Calculate wOBA
  woba_sum <- sum(filtered_data$woba_value, na.rm = TRUE)
  woba_denominator <- sum(filtered_data$events %in% at_bats_vector) +
    walk_count +
    hit_by_pitch_count +
    sum(filtered_data$events == 'sac_fly')
  woba <- woba_sum / woba_denominator
  
  #Calculate Barrel Rate (Barrel)
  barrel_rate <- barrel_count / PA
  
   # Calculate BB%
  walk_rate <- walk_count / (PA - hit_by_pitch_count)
  
  # Calculate K%
  strikeout_rate <- strikeout_count / PA
  
   # Calculate Z-Swing %
  
  z_swing <- sum(filtered_data$description %in% swings_vector & 
                 filtered_data$true_zone == 'strike', na.rm = TRUE) /
             sum(filtered_data$description %in% swings_vector)
  
  # Calculate O-Swing %
  
  o_swing <- sum(filtered_data$description %in% swings_vector & 
                 filtered_data$true_zone == 'ball', na.rm = TRUE) /
             sum(filtered_data$description %in% swings_vector)
  
  # Calculate ZO-Swing %
  
  zo_swing <- z_swing - o_swing
  
  # Calculate AEV
  average_ev <- mean(filtered_data$launch_speed, na.rm = TRUE)
  
  # Create a row for the current player in Metrics
  new_row <- data.frame(year = year,
                        player_id = player_id, 
                        PA = PA, 
                        AVG = batting_avg, 
                        OBP = on_base_percentage, 
                        SLG = slug_percentage, 
                        OPS = ops,
                        HR = hr,
                        wOBA = woba, 
                        Barrel = barrel_rate,
                        BB = walk_rate, 
                        K = strikeout_rate,
                        ZOsw = zo_swing,
                        AEV = average_ev)
                    
  # Append the row to the Metrics data frame
  Metrics <- rbind(Metrics, new_row)
  
  Metrics <- Metrics %>%
    filter(PA > 100)
  
}

# Create player info table and join

mlb_stats <- mlb_stats(stat_type = 'season', player_pool = 'All', stat_group = 'hitting',
                           season = year)

player_info <- mlb_stats %>%
  select(player_id, player_full_name, position_abbreviation,
         team_name, stolen_bases, caught_stealing)

Metrics <- Metrics %>%
  left_join(player_info, by = 'player_id') %>%
  select(player_id, player_full_name, position_abbreviation,
         team_name, everything(), stolen_bases, caught_stealing)

Metrics <- Metrics %>%
    rename(Name = player_full_name,
           Team = team_name,
           POS = position_abbreviation,
           SB = stolen_bases,
           CS = caught_stealing)

return(Metrics)
}
```

apply function to each year
```{r}
stats_22 <- get_stats(2022)
stats_21 <- get_stats(2021)
stats_20 <- get_stats(2020)
stats_19 <- get_stats(2019)
stats_18 <- get_stats(2018)

head(stats_18)
```

Edit 2020 data to estimate 162 game stats
```{r}
stats_20$PA <- stats_20$PA * 162 / 60
stats_20$HR <- stats_20$HR * 162 / 60
stats_20$SB <- stats_20$SB * 162 / 60
stats_20$CS <- stats_20$CS * 162 / 60
```

Create combined stats table
```{r}
combined_stats <- rbind(stats_22, stats_21, stats_20, stats_19, stats_18)

head(arrange(combined_stats, Name), 15)
```

Combine stats and contracts into one table
partition the data into train and validation data
```{r}
position_data <- function(combined_stats, contracts) {
  data <- contracts %>%
    left_join(combined_stats, by = c("name" = "Name", "year1" = "year")) %>%
  na.omit
  
  # Partition into train and valid
  
  RNGkind(sample.kind = 'Rounding')
  set.seed(42)
  train_rows <- sample(1:dim(data)[1], dim(data)[1]*.67)
  train <- data[train_rows,]
  valid <- data[-train_rows,]

result_list <- list(data = data,
                    train = train,
                    valid = valid)

return(result_list)
}
```

Apply to every position
```{r}
contracts_c <- get_contracts('catcher')
data_c <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_c)$data
train_c <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_c)$train
valid_c <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_c)$valid


contracts_1b <- get_contracts('1st-base')
data_1b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_1b)$data
train_1b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_1b)$train
valid_1b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_1b)$valid


contracts_2b <- get_contracts('2nd-base')
data_2b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_2b)$data
train_2b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_2b)$train
valid_2b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_2b)$valid


contracts_3b <- get_contracts('3rd-base')
data_3b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_3b)$data
train_3b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_3b)$train
valid_3b <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_3b)$valid


contracts_ss <- get_contracts('shortstop')
data_ss <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_ss)$data
train_ss <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_ss)$train
valid_ss <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_ss)$valid


contracts_of <- get_contracts('outfield')
data_of <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_of)$data
train_of <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_of)$train
valid_of <- position_data(combined_stats = combined_stats, 
                         contracts = contracts_of)$valid

```
# Pitchers

```{r}
get_stats_pitcher <- function(year) {
# From Statcast data, calculate metrics for qualified players
Statcast <- get(paste0('Statcast',year))
# Get unique values from the 'batter' column of Statcast
unique_playerID <- unique(Statcast$pitcher)

# Create empty data frame for Metrics
Metrics <- data.frame()
# Define label vectors and counts
hits_vector <- c('single', 'double', 'triple', 'home_run')
on_base_vector <- c(hits_vector, 'walk', 'hit_by_pitch','catcher_interf')
at_bats_vector <- c('field_out', 'strikeout', 'grounded_into_double_play', 'fielders_choice', 'fielders_choice_out', 'triple_play', 'strikeout_double_play', 'double_play', 'field_error', 'force_out', hits_vector)
plate_appearances_vector <- c(at_bats_vector, 'sac_fly', 'walk', 'hit_by_pitch', 'sac_bunt', 'sac_bunt_double_play', 'sac_fly_double_play', 'catcher_interf')


for (player_id in unique_playerID) {

  filtered_data <- filter(Statcast, pitcher == player_id & game_type == 'R')
  
  # Define count labels
  walk_count <- sum(filtered_data$events == 'walk')
  hit_by_pitch_count <- sum(filtered_data$events == 'hit_by_pitch')
  strikeout_count <- sum(filtered_data$events == 'strikeout')
  barrel_count <- sum(filtered_data$barrel, na.rm = TRUE)
  home_run_count <- sum(filtered_data$events == 'home_run')

  # Calculate Batters Faced
  BF <- sum(filtered_data$events %in% plate_appearances_vector)
  
  # Calculate wOBA-against
  woba_sum <- sum(filtered_data$woba_value, na.rm = TRUE)
  woba_denominator <- sum(filtered_data$events %in% at_bats_vector) +
    walk_count +
    hit_by_pitch_count +
    sum(filtered_data$events == 'sac_fly')
  woba_against <- woba_sum / woba_denominator
          
  
  # Calculate K%
  k_rate <- strikeout_count / BF
  
  # Calculate BB%
  bb_rate <- walk_count / BF
  
    # Calculate HR%
  hr_percentage <- home_run_count / BF
  
  # Barrel Rate - against
  barrel_against <- barrel_count / BF
  
  
  
  # Create a new row for the pitcher with temp columns for FIP
  new_row <- data.frame(year = year,
                        player_id = player_id,
                        BF = BF,
                        wOBA = woba_against,
                        K = k_rate,
                        BB = bb_rate,
                        HR = hr_percentage,
                        Barrel = barrel_against,
                        tHR = home_run_count,
                        tBB = walk_count,
                        tHBP = hit_by_pitch_count,
                        tK = strikeout_count)
  
 # Append the row to the Metrics data frame
  Metrics <- rbind(Metrics, new_row)
  
  Metrics <- Metrics %>%
    filter(BF > 100)
                        
}

# Create player info table and join

mlb_stats <- mlb_stats(stat_type = 'season', player_pool = 'All', stat_group = 'pitching',
                           season = year)

player_info <- mlb_stats %>%
  select(player_id, player_full_name,
         era, innings_pitched, wins, saves, whip)

Metrics <- Metrics %>%
  left_join(player_info, by = 'player_id') %>%
  select(player_id, player_full_name, everything(), era, innings_pitched, wins, saves, whip)

# Convert player_info stats to numerics
Metrics$era <- as.numeric(Metrics$era)
Metrics$innings_pitched <- as.numeric(Metrics$innings_pitched)
Metrics$whip <- as.numeric(Metrics$whip)

# Further IP operation
Metrics$IP <- if_else(Metrics$innings_pitched %% 1 == 0,
                              Metrics$innings_pitched, 
                            if_else(Metrics$innings_pitched %% 1 == 1,
                                Metrics$innings_pitched + 0.23,
                                Metrics$innings_pitched + 0.47))


 # Calculate FIP
  Metrics$FIP <- ((13*Metrics$tHR + 3*(Metrics$tBB+Metrics$tHBP) - 2*Metrics$tK) /
                      Metrics$innings_pitched) + 3.2
  Metrics <- Metrics %>%
    select(-c(tHR, tBB, tK, tHBP, innings_pitched))
          

Metrics <- Metrics %>%
    rename(Name = player_full_name,
           ERA = era,
           Wins = wins,
           Saves = saves,
           WHIP = whip)

return(Metrics)
}
```



apply function to each year
```{r}
stats_22_pitcher <- get_stats_pitcher(2022)
stats_21_pitcher <- get_stats_pitcher(2021)
stats_20_pitcher <- get_stats_pitcher(2020)
stats_19_pitcher <- get_stats_pitcher(2019)
stats_18_pitcher <- get_stats_pitcher(2018)

head(stats_18_pitcher)
```

Edit 2020 data to estimate 162 game stats
```{r}
stats_20_pitcher$BF <- stats_20_pitcher$BF * 162 / 60
stats_20_pitcher$Wins <- stats_20_pitcher$Wins * 162 / 60
stats_20_pitcher$Saves <- stats_20_pitcher$Saves * 162 / 60
stats_20_pitcher$IP <- stats_20_pitcher$IP * 162 / 60
```

Create combined stats table
```{r}
combined_stats_pitcher <- rbind(stats_22_pitcher, stats_21_pitcher, stats_20_pitcher,
                        stats_19_pitcher, stats_18_pitcher)

head(arrange(combined_stats_pitcher, Name), 15)
```

Combine stats and contracts into one table
partition the data into train and validation data
```{r}
pitcher_data <- function(combined_stats, contracts) {
  data <- contracts %>%
    left_join(combined_stats, by = c("name" = "Name", "year1" = "year"),
              suffix = c('.1','.2')) %>%
    left_join(combined_stats, by = c("name" = "Name", "year2" = "year")) %>%
    left_join(combined_stats, by = c("name" = "Name", "year3" = "year")) %>%
    filter(rowSums(is.na(.)) < 20)

  
  # Partition into train and valid
  
  RNGkind(sample.kind = 'Rounding')
  set.seed(42)
  train_rows <- sample(1:dim(data)[1], dim(data)[1]*.67)
  train <- data[train_rows,]
  valid <- data[-train_rows,]

result_list <- list(data = data,
                    train = train,
                    valid = valid)

return(result_list)
}
```

Apply to every position
```{r warning=FALSE}
contracts_sp <- get_contracts('starting-pitcher')
data_sp <- pitcher_data(combined_stats = combined_stats_pitcher, 
                         contracts = contracts_sp)$data
train_sp <- pitcher_data(combined_stats = combined_stats_pitcher, 
                         contracts = contracts_sp)$train
valid_sp <- pitcher_data(combined_stats = combined_stats_pitcher, 
                         contracts = contracts_sp)$valid

contracts_rp <- get_contracts('relief-pitcher')
data_rp <- pitcher_data(combined_stats = combined_stats_pitcher, 
                         contracts = contracts_rp)$data
train_rp <- pitcher_data(combined_stats = combined_stats_pitcher, 
                         contracts = contracts_rp)$train
valid_rp <- pitcher_data(combined_stats = combined_stats_pitcher, 
                         contracts = contracts_rp)$valid


```











